---
title: "Plotting Electric Vehicle Open Data"
subtitle: "with ggplot2 and friends"
date: "2025-06-10"
---

## Goal

Today we're going to:

- Introduce plotting with R and the `ggplot2` package
- Provide resources for further reading
- Highlight some best practices


## The Setup

I'll be using `renv` - however, `renv` is not the
focus of today. Plotting in R using the `ggplot2` package is.
Be sure to check out the doco [here](https://rstudio.github.io/renv/).
We're going to start by attaching the libraries with `library()`.

```{r setup}
# suppressing the startup messages
library(tidyverse) |> suppressPackageStartupMessages()

# ggplot2 is a core tidyverse package so it's
# included in the namespace when folks `library(tidyverse)`
```


## The Data

Today we're going to use the Electric Vehicle data from our #tidytuesday
community [challenge](https://github.com/CTOpenData/r-user-group/tree/d748587f4802578325b6a9d55c59b71563ba260d/tidy-tuesday).

### Grabbing the Data

Unfortunately, `RSocrata` was pulled from the CRAN - so we're going to have to
homebrew our own solution.

Fortunately, I put together a gist on github. You may recall Charles Powell
introduced these last year. You can take a look at the gist
[here](https://gist.github.com/asenetcky/52bf3fa10a2dff08f62da96e2347e018).

We can use `devtools` to source the gist and grab the function I wrote
to download *all* the EV data.

::: {.callout-tip}
  `devtools` is a collection of packages, like the `tidyverse`
  that aims to smmoth over alot of *rough edges* in development.
  Primarily around building R packages or interacting with **GitHub**,
  but there is all sorts of neat things in `devtools` - so be sure
  to check it out.
:::

```{r}
#| label: gist

devtools::source_gist(
  id = "https://gist.github.com/asenetcky/52bf3fa10a2dff08f62da96e2347e018",
  sha1 = "5b27e785462239543a6cda235382ebec7c381471"
)

# function exist?
head(pull_odp)
```

We have our function, now lets grab that open data.

```{r}
#| label: data-grab
ev <- pull_odp(resource = "y7ky-5wcz") # this is the ODP "four by four"
```

### Data Recon

Lets take a look at this dataset so we can have a better idea
of what we are working with.

```{r}
head(ev)
```

```{r}
glimpse(ev)
```

So we have `20` columns and `60,489` rows. 

### Quick Wrangle!

I don't want to spoil anyone's sense of discovery with these data.
So I am going to just quickly wrangle it and skip over it,
but feel free to take a look at the source code when you have time.
Or after you're done experimenting with it.

```{r}
#| label: wrangle
#| code-fold: true
#| code-summary: "Show the code"

data <-
  ev |>
  mutate(
    across(
      .cols = c(
        id,
        vehicleweight,
        vehicleyear,
        vehicledeclaredgrossweight,
        vehiclerecordedgvwr,
      ),
      .fns = as.numeric
    ),
    across(
      .cols = c(
        registration_date_start,
        registration_date_expiration
      ),
      .fns = as_date
    )
  )

char_data <-
  data |>
  select(where(is.character))

char_data <-
  data |>
  select(where(is.character))

maybe_factors <-
  map(
    char_data,
    \(var) {
      count(char_data, {{ var }}, sort = TRUE)
    }
  ) |>
  keep(\(df) nrow(df) < 100) |>
  names()

ev <-
  data |>
  mutate(
    across(
      .cols = all_of(maybe_factors),
      .fns = \(col) {
        col |>
          forcats::as_factor() |>
          forcats::fct_infreq()
      }
    )
  )

```


## `ggplot2` and friends

`ggplot2` is now over 10 years old and a stable and trusted package
for visualizations.  It is based on the core philosphy around
Leland Wilkinson's _The Grammar of Graphics_.  

> _Wilkinson, L. (2005), The Grammar of Graphics, 2nd ed., Springer._

If you like, you can read up on how `ggplot2` incorporates
those ideals [ggplot2: Elegant Graphics for Data Analysis (3e)](https://ggplot2-book.org/).

It's probably safe to say it is a cornerstone of the R community.  
The examples, and [documentation](https://ggplot2.tidyverse.org/) 
you can find online is often extensive and well written. 
`python` even has a port of it with the `plotnine` package.

### Syntax Changes: a note about `+`

Right away folks are going to notice that `ggplot2` uses
a different syntax than the usual base pipe `|>` or 
maggritr pipe `%>%` that you will often see out in
the wild.  This is primarily down to how old the 
package is, however, I find that `+` more accurately
describes the thought process involved with crafting
a plot.

With pipes, action happen linearly in order, whereas
you an think of `ggplot2` plots like building up a 
layer cake. You are _adding_ layers, quite literally
with a `+`. 


## ggplot 2 in action

`ggplot2` is fairly flexible in how you structure your
commands, so you're likley to see some variations between
developers.

```{r}
#| labek: setting-the-stage

# we can start with the data, or a call to `ggplot()`
# I prefer starting with the data and piping it in, otherwise
# folks can use ggplot(data = ev, ...)

ev |>
  ggplot(
    mapping = aes(
      # we're mapping our aesthetic, think of it like a base we pin layers to
      x = registration_date_start, # x axis
      y = vehicletype # y axis
    )
  )

```

If we just run the above, we'll get a blank plot. However, take a quick look at 
that - so much of the stage has already been laid out for us.  `ggplot2` has fairly
sensible defaults, so for the folks who want to get in and get out quickly, they
can do that.  This allows plotting and visualizations to be used as one part
of an exploratory analysis, and not just as a final, finished product, because
it is so quick and easy to do once users are acquainted with `ggplot2`.

### Adding Details

How about we fill in some useful details, no?

```{r}
#| label: details

# Let's add some layers!
ev |>
  ggplot(
    mapping = aes(
      # we're mapping our aesthetic, think of it like a base we pin layers to
      x = registration_date_start # x axis
    )
  ) +
  geom_bar() # How about some bars?
```

Goodness there is alot going on there! 

Does day of week make a difference? Let's find out.


```{r}
#| label: day-of-week

ev |>
  select(registration_date_start) |>
  mutate(
    dow = wday(
      registration_date_start,
      label = TRUE
    )
  ) |>
  count(registration_date_start, dow) |>
  ggplot() +
  geom_boxplot(aes(dow, n))
```


Notice again that we had some fairly sensible defaults, without having to
dig into all the arguments of the functions.


### The Ask
Let's pretend our supervisor want's to compare the `vehicletype` over `time`.

Let's do another quick wrangle so we can take a look.

```{r}
agg_by_year_quarter_type <-
  ev |>
  mutate(
    year = year(registration_date_start),
    quarter = quarter(registration_date_start),
    vehicletype = forcats::fct_lump_prop(
      vehicletype,
      prop = .01
    )
  ) |>
  group_by(year, quarter, vehicletype) |>
  count(name = "type_count") |>
  ungroup()

```


```{r}
agg_by_year_quarter_type |>
  ggplot() +
  geom_col(
    aes(
      x = quarter,
      y = type_count
    )
  )
```

what about year and quarter?

```{r}
agg_by_year_quarter_type |>
  ggplot() +
  geom_col(
    aes(
      x = quarter,
      y = type_count
    )
  ) +
  facet_grid(cols = vars(year))
```

Didn't we want to compare `vehicletype`?
let's do that....but how?

Using color!

### Adding Color based on Features in the Data

```{r}
agg_by_year_quarter_type |>
  ggplot() +
  geom_col(
    aes(
      x = quarter,
      y = type_count,
      # now colors are added to break out based on values in the data
      fill = vehicletype
    ),
    position = position_dodge()
  ) +
  facet_grid(cols = vars(year))
```

Let's natural log transform these data to get rid of that skew.

```{r}
#| label: natural-log

# using lubridate and forcats
agg_by_year_quarter_type <-
  agg_by_year_quarter_type |>
  mutate(log_type = log(type_count))

head(agg_by_year_quarter_type)
```

And try again.

```{r}
agg_by_year_quarter_type |>
  ggplot() +
  geom_col(
    aes(quarter, log_type, fill = vehicletype),
    position = position_dodge()
  ) +
  facet_grid(cols = vars(year))
```


Now, I want to save some typing. `ggplot2` objects can be saved and combined
in fun and exciting ways.

```{r}
base <-
  agg_by_year_quarter_type |>
  ggplot(
    aes(quarter, log_type, fill = vehicletype)
  ) +
  facet_grid(cols = vars(year))

# the same thing again
base
col_base <- base + geom_col(position = position_dodge())
col_base
```

Cool... now we can focus on making this better.

### Tweaking the look and feel

```{r}
# there are many to choose from - and you can even build your own!

col_base + theme_classic()

col_base + theme_bw()

col_base + theme_minimal()

```

```{r}
# now for the labels

with_labels <-
  col_base +
  theme_classic() +
  # add some labels
  labs(
    title = "This is a Title",
    subtitle = paste("Subtitle as of", lubridate::today()),
    caption = "this is a caption - Open Data is Awesome!",
    tag = "This is a tag",
    # the next bit refers to the mapped variables x, y etc... BUT it may
    # not always be x and y
    x = "This is an x axis",
    y = "This is a y axis",
    fill = "my fill values"
  )

with_labels

```


### Some Polish

```{r}

base +
  geom_col(
    position = position_dodge(), # I want bars side by side, not stacked
    col = "black" # just a black outline
  ) +
  scale_fill_viridis_d() + # use viridis color scale
  theme_classic() + # we can use a default theme
  theme(
    # and add our own components!
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  labs(
    x = "", # I don't want axis labels
    y = "",
    fill = "",
    title = "Natural Log of Vehicle Type Count by Quarter by Year"
  )
```


